Discovery of s-process stars in the LAMOST survey

### 2018-01-25 article/ms.tex
foo
merge

@@

\author[B.~J. Norfolk et al.]{B.~J. Norfolk,$^{1}$\thanks{E-mail: bjee7@student.monash.edu (MU)}
A.~R. Casey,$^{1,2}$
M.~T. Miles,$^{1}$
A.~J. Kemp,$^{1}$ 
A.~I. Karakas,$^{1}$ \newauthor
J.~C. Lattanzio,$^{1}$
and K.~C. Schlaufman$^{3}$
\\
$^{1}$School of Physics \& Astronomy, Monash University, Clayton 3800, Victoria, Australia\\
$^{2}$Faculty of Information Technology, Monash University, Clayton 3800, Victoria, Australia\\
$^{3}$Department of Physics and Astronomy, Johns Hopkins University, 3400 N Charles St., Baltimore, MD 21218, USA
}
\date{Accepted XXX. Received YYY; in original form ZZZ}

@@

\begin{abstract}
Stars with peculiar enhancements of carbon and heavy elements ($Z > 30$) are historically referred to as `Barium stars'. While this chemical abundance pattern can be explained by thermally pulsing asymptotic giant branch (AGB) stars, many so-called Barium stars have not evolved through the red giant phase and therefore cannot be AGB stars. These less-evolved stars are generally thought to have accreted mass from a more-evolved AGB companion. For this reason the frequency and properties of less-evolved Barium stars are uniquely informative on the binary fraction, AGB yields, mass transfer, and stellar evolution. Here we present the discovery of 967 s-process candidates from 454,180 giant stars observed by LAMOST. Our sample includes; 257 Barium stars, 49 stars exhibiting only significant barium enhancement, and 661 exhibiting only strontium enhancement. This sample constitutes the largest number of s-process enhanced stars ever discovered, and represents the most comprehensive search for Barium stars in the Milky Way. All of our Barium stars are consistent with membership in the Milky Way disk, and we find an \todo{steady/increasing/decreasing} fraction of Barium stars with stellar metallicity, as \todo{expected} from AGB yields and stellar evolution theory. We find up to 51 Barium stars exhibit some evidence of technetium from LAMOST spectra, and most (46/51) have $\log{g} > 2$, which suggests they cannot be AGB stars and may indicate a recent ($100\,\textrm{Myr}$) mass accretion event. Most (178/257) Barium stars in our sample show strong carbon enhancement, consistent with the literature. However, despite previous claims indicating that Barium stars are preferentially enhanced in sodium due to the NeNa nucleosynthesis cycle occurring in massive AGB stars, only 5 ($<2$\,\%) of our Barium stars show any evidence of enhancement in sodium. \todo{A sentence about AGB yield comparison.} When binary properties become available from end-of-mission Gaia astrometry, this sample of s-process stars will be critical in explaining mass transfer in binary systems and refining AGB yields across a range of masses and metallicities.
\end{abstract}

@@

Barium stars as first recognised by \citet{Bidelman1951}, are chemically unique objects that exhibit envelopes with an overabundance of carbon and heavy elements (Z > 30) in comparison to the Sun. Barium stars exist as either intrinsic or extrinsic objects; intrinsic Barium stars are in the TP-AGB (thermally pulsing-asymptotic giant branch) phase, where the overabundance of both carbon and heavy elements are produced through the s-process and exist on the surface as the result of a thermal pulsing cycle. According to the mass-transfer hypothesis, extrinsic Barium stars are a consequence of stellar wind accretion \citep{boffin1988,jorissen1992} or Roche-lobe overflow \citep{webbink1986}, and are within a binary system containing a previous TP-AGB companion star (intrinsic Barium star) in its final phase as a white dwarf \citep{bohm1980,bohm1984}. In fact, \citet{mcclure1983} determines 85\% of all Barium stars are in binary systems and claims that those that appear singular are actually pole-on, or highly eccentric binaries with significant radial velocity variations only occurring in a small phase range as detailed by \citet{pourbaix2004}. For these reasons, the properties and occurrence rate of Barium stars are informative of the theoretical birth rate of AGB stars, the binary star fraction as a function of metallicity, the mass ratio of binary stars, as well as AGB yields across different masses and metallicities.

The slow neutron capture process (s-process), occurring in the interior of AGB stars, synthesises roughly half of all elements heavier than iron \citep[e.g.][]{busso1999,travaglio2001,herwig2005,romano2010,kobayashi2011,prantzos2012,bisterzo2014,karakas12016}. Late in the AGB phase, during the thermally pulsing-AGB (TP-AGB) phase, thermal instabilities occur in the He shell every $10^5$ years or so, depending on the H-exhausted core mass. These energy bursts drive a convective zone that sweeps the entire region lying between the core mass and He-shell; mixing the products of nucleosynthesis within these regions. The energy from the thermal pulses force the star to expand, pushing the H-shell out to cooler regions and allowing the convective envelope to move inwards to regions previously mixed by the thermal pulse driven convective zones. This expansion and resulting inward movement is described as the third dredge up (TDU), and is theorised to occur after each thermal pulse. During the TP-AGB phase, enrichment on the surface in $^{12}$C and heavy elements produced by the s-process is a result of the repeated downwash extensions of TDU \citep[e.g.][]{busso2001}. The star contracts PDU and reignites the H-shell producing the majority of the surface luminosity for the next interpulse period. The interpulse, thermal pulse, and dredge up cycle may occur numerous times and is dependent on the initial mass, composition, and mass-loss rate of the star.

Barium giants form within a metallicity-dependent initial mass range (approximately $0.8 - 8\,M_{\odot}$), with the minimum mass for core helium and carbon burning decreasing as metallicity decreases. The age of these stars vary considerably, with observable stars in metal-poor globular clusters hosting populations of $\approx$12Gyr. In contrast, metal rich, young observable stars may only reach $\approx$100Myr, this includes stars that are at the core carbon burning limit or very close to it \citep[e.g.][]{whitelock2013}.

@@

In this paper we analyse 454,180 giant stars from the second LAMOST data release (DR) to identify and examine 967 s-process candidates including Barium stars, and exclusively enhanced barium and strontium stars, respectively. In Section \ref{sec:methods} we describe the observations and candidate selection, as well as the analysis of high-resolution follow-up observations obtained for a few candidates. In Section \ref{sec:dis} we discuss the properties of our Barium stars in context of existing literature. We provide concluding remarks in Section \ref{sec:con}

@@

\subsubsection{Candidate selection} \label{sec:cand}
The Barium stars analysed in this research were obtained by filtering for significant flux residuals, taken as, a disparity between the normalised LAMOST flux and the data driven model derived by \citet{ho2017}. Specifically, we identified Barium stars by searching for \ion{Ba}{II} and \ion{Sr}{II} enhancement at the 4554\,\AA\ and 4077\,\AA, respectively. A negative residual, as shown in a candidate Barium star in Figure \ref{fig:figure1}, illustrates an enhancement in \ion{Ba}{II} and \ion{Sr}{II} at lines 4554\,\AA\, 4077\,\AA. These \ion{Ba}{II} and \ion{Sr}{II} resonance lines are very strong, allowing us to identify stars enhanced in neutron-capture elements given a well-fit model for the data. Stars identified to just show high levels of barium required enhancement at both the 4554\,\AA\ and 4934\,\AA\ resonance lines, and for stars just enhanced in strontium we looked for significant residuals at the 4554\,\AA\ and 4215\,\AA\ absorption lines.

We used five filters that each spectrum, for each set of absorption lines, had to meet in order to be considered a s-process candidate; these included:

@@

\begin{enumerate} 
\item Profile amplitude for both enhancement lines must be $A < -0.05$.
\item Both amplitudes must be measured within 3$\sigma$ ($|A|/\sigma _A$ < 3).
\item The wavelength at each absorption line must be within 2\,\AA\ ($\lambda$ < 2).
\item The reduced $\chi^2$ from \emph{The Cannon} must be $\chi_r^2 < 3$.
\item And finally, the LAMOST spectra must have a signal-to-noise ratio of $S/N > 30\,\textrm{pixel}^{-1}$.
\end{enumerate}
In addition, a visual inspection was undertaken to exclude any results containing, false positives, candidates exhibiting data reduction issues, apparent absorption finer then the spectral resolution, and overly noisy normalised LAMOST spectra. This combined method generated 967 s-process candidates including; 257 Barium stars, 49 stars exhibiting exclusively barium enhancement, and 661 exhibiting exclusively strontium enhancement, as listed in Table \ref{table:table1}.

@@

\subsubsection{Enhancements due to sodium, technetium, and carbon}
We performed an identical analysis to the process described in Section \ref{sec:cand} in order to identify sodium, technetium, or carbon enhancement in our sample of 257 Barium stars. We used the sodium doublet lines at 5889\,\AA\ and 5895\,\AA\ to select Barium stars enhanced in sodium, and required significant absorption in both features. Only 5/257 Barium stars met this criteria,  \todo{however something about the dust for those stars}. For technetium enhancement we searched for significant residual deviations at the 4049\,\AA, 4238\,\AA, 4262\,\AA, 4297\,\AA, and 5924\,\AA. These absorption lines are extremely weak, and would require a substantial amount of technetium before it would be visible in a high S/N LAMOST spectrum. We found 51 Barium stars exhibited some level of significant enhancement (based on the filters above) at a single technetium line, yet none showed enhancement at more than one technetium line. We note that the single enhancement near technetium for these 51 stars may be the consequence of data reduction/calibration artefacts, but we include details of which stars matched on technetium in Table \ref{tab:table} nonetheless.
Finally, for carbon enhancement we searched for significant deviations at the CH and G band near 4300\,\AA. We found 178/257 stars to show significant carbon enhancement (e.g., $[\textrm{C/M}] \gtrsim 0.5$).

@@

\begin{figure}
	% To include a figure from a file named example.*
	% Allowable file formats are eps or ps if compiling using latex
	% or pdf, png, jpg if compiling using pdflatex
	\includegraphics[width=\columnwidth]{HRdummy.png}
    \caption{HR Diagram}
    \label{fig:figure2}
\end{figure}

>>>>

% Please keep new commands to a minimum, and use \newcommand not \def to avoid
% overwriting existing commands. Example:
%\newcommand{\pcm}{\,cm$^{-2}$}	% per cm-squared

@@

% The list of authors, and the short list which is used in the headers.
% If you need two or more lines of authors, add an extra line using \newauthor
\author[B.J. Norfolk et al.]{B.J. Norfolk,$^{1}$\thanks{E-mail: bjee7@student.monash.edu (MU)}
A.R. Casey,$^{1,2}$
M.T. Miles,$^{1}$
A.J. Kemp,$^{1}$ 
A.I. Karakas,$^{1}$ \newauthor
J.C. Lattanzio,$^{1}$
and K.C. Schlaufman 
\\
$^{1}$School of Physics \& Astronomy, Monash University, Clayton 3800, Victoria, Australia\\
$^{2}$Faculty of Information Technology, Monash University, Clayton 3800, Victoria, Australia\\
}
\date{Accepted XXX. Received YYY; in original form ZZZ}

@@

\begin{abstract}
Stars with peculiar enhancements of carbon and heavy elements ($Z > 30$) are historically referred to as Barium stars. While this chemical abundance pattern can be explained by thermally pulsing asymptotic giant branch (AGB) stars, many Barium stars have not evolved through the red giant phase and therefore cannot be AGB stars. These stars are generally thought to have accreted mass from a more evolved AGB companion. For this reason the frequency and properties of less-evolved Barium stars is uniquely informative on the binary fraction, AGB yields, mass transfer, and stellar evolution. Here we present the discovery of 967 s-process candidates including; 257 Barium stars, 49 stars exhibiting exclusively barium enhancement, and 661 exhibiting exclusively strontium enhancement from 454,180 giant stars observed by LAMOST. This sample constitutes the largest sample of s-process enhanced stars discovered to date, and the most comprehensive search for Barium stars in the Milky Way. \todo{Nearly all} of our Barium stars are disk stars, and we find an \todo{steady/increasing/decreasing} fraction of Barium stars with stellar metallicity, as \todo{expected} from AGB yields and stellar evolution theory. 51 Barium stars exhibited tecnetium enhancement in a single Tc emission line suggesting the exsistence of AGB stars in our sample. Most (178/257) of our sample show strong carbon enhancement consistent with the literature. Despite previous claims indicating that Barium stars are preferentially enhanced in sodium due to the NeNa nucleosynthesis cycle occurring in \todo{massive} AGB stars, only 5 ($<2$\,\%) of our Barium stars show any evidence of enhancement in sodium. \todo{This implies ....}
\end{abstract}

@@

Barium stars as first recognized by \citet{Bidelman1951}, are chemically unique objects that exhibit envelopes with an overabundance of carbon and heavy elements (Z > 30) in comparison to the Sun. Barium stars exist as either intrinsic or extrinsic objects; intrinsic Barium stars are in the TP-AGB (thermally pulsing-asymptotic giant branch) phase, where the overabundance of both carbon and heavy elements are produced through the s-process and exist on the surface as the result of a thermal pulsing cycle. According to the mass-transfer hypothesis, extrinsic Barium stars are a consequence of stellar wind accretion \citep{boffin1988,jorissen1992} or Roche-lobe overflow \citep{webbink1986}, and are within a binary system containing a previous TP-AGB companion star (intrinsic Barium star) in its final phase as a white dwarf \citep{bohm1980,bohm1984}. In fact, \citet{mcclure1983} determines 85\% of all Barium stars are in binary systems and claims that those that appear singular are actually pole-on, or highly eccentric binaries with significant radial velocity variations only occuring in a small phase range as detailed by \citet{pourbaix2004}. For these reasons, the properties and occurrence rate of Barium stars are informative of the theoretical birth rate of AGB stars, the binary star fraction as a function of metallicity, the mass ratio of binary stars, as well as AGB yields across different masses and metallicities.

The slow neutron capture process (s-process), occuring in the interior of AGB stars, synthesizes roughly half of all elements heavier than iron \citep[e.g.][]{busso1999,travaglio2001,herwig2005,romano2010,kobayashi2011,prantzos2012,bisterzo2014,karakas12016}. Late in the AGB phase, during the thermally pulsing-AGB (TP-AGB) phase, thermal instabilities occur in the He shell every $10^5$ years or so, depending on the H-exhausted core mass. These energy bursts drive a convective zone that sweeps the entire region lying between the core mass and He-shell; mixing the products of nucleosynthesis within these regions. The energy from the thermal pulses force the star to expand, pushing the H-shell out to cooler regions and allowing the convective envelope to move inwards to regions previously mixed by the thermal pulse driven convective zones. This expansion and resultsing inward movement is classified as the third dredge up (TDU), and is theorised to occur after each thermal pulse. During the TP-AGB phase, enrichent on the surface in $^{12}$C and heavy elements produced by the s-process is a result of the repeated downwash extensions of TDU \citep[e.g.][]{busso2001}. The star contracts PDU and reignites the H-shell producing the majority of the surface luminosity for the next interpulse period. The interpulse, thermal pulse, and dredge up cycle may occur numerous times and is dependent on the intitial mass, composition, and mass-loss rate of the star.

Barium giants form within a metallicity-dependent initial mass range (approximately 0.8 - 8$M_{\odot}$), with the minimum mass for core helium and carbon burning decreasing as metallicity decreases. The age of these stars vary considerably, with observable stars in metal-poor globular clusters hosting populations of $\approx$12Gyr. In contrast, metal rich, young observable stars may only reach $\approx$100Myr, this includes stars that are at the core carbon burning limit or very close to it \citep[e.g.][]{whitelock2013}.

@@

In this paper we analyse 454,180 giant stars from LAMOST data release (DR) 2 to identify and examine 967 s-process candidates including Barium stars, and exclusvely enhanced barium and strontium stars, respectively. Specifcally, the Barium star giants studied here display especially strong s-process enhancement for \ion{Ba}{II} and \ion{Sr}{II} at lines 4554\AA \hspace{0.2mm} and 4077\AA, respectively. Our exclusive barium and strontium candidates show s-process enhancement in 4554\AA \hspace{0.2mm} and 4934\AA for barium enhancement, and, 4554\AA \hspace{0.2mm} and 4215\AA for strontium enhancement. In Section \ref{sec:methods} we describe the observations and candidate selection, as well as the analysis follow up our follow-up observing campaign with Magellan. In Section \ref{sec:dis} we discuss the properties of our Barium stars in context of existing literature. We provide concluding remarks in Section \ref{sec:con}

@@

\subsubsection{Candidate selection} \label{sec:cand}
The Barium stars analysed in this research were obtained by filtering for significant flux residuals, taken as, a disparity between the normalised LAMOST flux and the data driven model derived by \citet{ho2017}. A negative residual as shown in the randomly selected Figure \ref{fig:figure1}, illustrates an enhancement in \ion{Ba}{II} and \ion{Sr}{II} at lines 4554\AA \hspace{0.2mm}, 4077\AA. These \ion{Ba}{II} and \ion{Sr}{II} resonance lines are very strong, allowing us to identify stars enhanced in neutron-capture elements given a well-fit model for the data. We used five filters that each spectrum, for both enhancement lines, had to meet in order to be considered a s-process candidate; these included:

@@

\begin{enumerate} 
\item Profile amplitude for both enhancement lines must exceed A < -0.05.
\item Both amplittudes must be measured within 3$\sigma$ ($|A|/\sigma _A$ < 3).
\item The wavelength at each absorption line must be within $2$\AA \hspace{0.2mm}($\lambda$ < 2).
\item Reduced $\chi^2$ from The Cannon must be < 3.
\item And, the LAMOST spectra must have a signal-to-noise ratio of S/N > 30 per pixel.
\end{enumerate}
In addition, a visual inspection was undertaken to exclude any results containing, false positives, candidates exhibiting data reduction issues, apparent absorption finer then the spectral resolution, and overly noisy normalised LAMOST spectra. This combined method generated 967 s-process candidates including; 257 Barium stars, 49 stars exhibiting exclusively barium enhancement, and 661 exhibiting exclusively strontium enhancement, as listed in Table \ref{table:table1}.

@@

\subsubsection{Further enhancements}
We undertook an identical analysis to the process described in Section \ref{sec:cand} for sodium, technetium, and carbon enhancement. We filtered for negative residuals in the enhanced sodium doublet lines 5889.950\AA\hspace{0.2mm} and 5895.924\AA. Our analysis produced 5 matches with our Barium star candidates. 51 canditdates were discovered to be enhanced in a single technetium line. And, for the CH and G band region at approximately 4300\AA\hspace{0.2mm} we matched 178 enhanced stars out of our total 257 stars.

@@

\begin{figure}
	% To include a figure from a file named example.*
	% Allowable file formats are eps or ps if compiling using latex
	% or pdf, png, jpg if compiling using pdflatex
	\includegraphics[width=\columnwidth]{HRdummy.png}
    \caption{effective temperature $T_\rm{eff}$ and surface gravity $\log{g}$ for 257 candidate Barium stars from LAMOST. Stars with $\log{g} > 2$ (as marked) have not evolved through the thermally pulsing asymptotic giant branch phase, and therefore their abundance signature must be the result of an extrinsic pollution event. The classification of stars with $\log{g} < 2$ would require additional data}
    \label{fig:figure2}
\end{figure}
